A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-10-22 paper.tex
additional discussion in dust edge instability
revising with youdin's comments
edits in response to youdin's comments
rough edits for youdin's comments
head to toe edits

\usepackage{amsmath}
\usepackage{paralist}
\usepackage{url}
\usepackage{bm}
%\usepackage{hyperref}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{\tilde{\epsilon}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is analogous to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty accretion disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show that in the limit of perfect coupling, dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
% 
We also extend 
the vertical shear instability (VSI) to locally isothermal, dusty 
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilize modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We discuss thermodynamical interpretations of 
dust-gas drag instabilities, for example the streaming
instability in protoplanetary disks, as a phase lag between the pressure 
and density of a dusty gas, much like pulsational instabilities 
in stars.  
\end{abstract}

@@

The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for the two-fluid and Lagragian
particles framework; but is straight forward to simulate in the
one-fluid formulation \citep{price15}.  The single fluid formalism
also has the important advantage that it is 
amenable to standard analytical methods
\citep{youdin05a,jacquet11}. This is the necessary  approach to gain
physical insight and explain some of the above newer dusty
instabilities.

In this work, we are motivated to develop a general analytical
framework to  study the stability of dusty protoplanetary
disks by drawing on the physical and mathematical parallel between dusty gas and standard
hydrodynamics. This is possible in certain limits of astrophysical
interest.

@@

Under these approximations, we show that the one-fluid equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is is analogous to the advection
of entropy following an adiabatic fluid. We will formally relate the 
entropy of isothermal dusty gas to its dust content. Finite  
dust-gas drag can be interpreted as heating/cooling: the relative
motion between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. We explicitly demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We 
show that strictly isothermal protoplanetary disks are generally
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations.

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations in disguise. 
We describe approximate  equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty disk in various
limits. There, we give a thermodynamical interpretation of
instabilities caused by directly by dust-gas drag. In
\S\ref{linear_problem} we list the linearized equations, which we
solve numerically in \S\ref{results} to study the dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

>>>>

\usepackage{amsmath}
\usepackage{paralist}
\usepackage{url}
\usepackage{bm}
%\usepackage{hyperref}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\rich}{\mathrm{Ri}}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{f_\mathrm{d}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}}
\newcommand{\seff}{S_\mathrm{eff}}
\newcommand{\tcool}{t_\mathrm{cool}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

@@

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is analogous to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty accretion disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show that in the limit of perfect coupling, dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
However, sufficiently sharp radial gradients in the dust-to-gas ratio
can destabilize the disk.  
We also extend 
the vertical shear instability (VSI) to locally isothermal, dusty 
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilize modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We also duscuss thermodynamical interpretations of 
dust-gas drag instabilities, and show that the streaming
instability in protoplanetary disks occur because drag causes 
a phase lag between the gas pressure  
and dust density, much like pulsational instabilities 
in stars.  
\end{abstract}

@@

The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for two-fluid models and 
following Lagragian particles; but is straight forward to simulate in the 
one-fluid formulation \citep{price15}.  The single fluid formalism
also has the important advantage that it is 
amenable to standard analytical methods \citep{youdin05a,jacquet11}.  
This is the necessary  approach to gain physical insight.

In this work, we are motivated to develop a general analytical
framework to study the stability of dusty protoplanetary 
disks by drawing on the physical and mathematical parallel between
dusty gas and standard hydrodynamics. This is possible in certain limits of astrophysical
interest, allowing us to interpret dusty phenomenon through standard
hydrodynamics.

@@

Under these approximations, we show that the one-fluid equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is is analogous to the advection
of entropy following an adiabatic fluid. We will formally relate the 
effective entropy of isothermal dusty gas to its dust content. Finite  
dust-gas drag act as effective heating/cooling: the relative
motion between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. We explicitly demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We 
show that strictly isothermal protoplanetary disks are generally
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We put forward a thermodynamical thermodynamical interpretation of
dust-gas drag instabilities, and apply it to the
streaming instability \citep[SI, ][]{youdin05a,youdin07b,jacquet11}. 
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}.

This paper is organized as follows. In \S\ref{setup} we list the
one-fluid equations describing tightly-coupled dusty gas. We show that for a
fixed equation of state, these are in 
fact the usual hydrodynamic equations in disguise. 
We describe equilibrium solutions in \S\ref{eqm} which are used in
explicit calculations. In \S\ref{limits} we discuss
general stability properties of dusty disks. We then move onto
explicit stablity analyses. In  
\S\ref{linear_problem} we list the linearized equations, and solve 
them for the SI in \S\ref{si}, and in 
 \S\ref{results} study dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

### 2016-10-22 summary.tex
edits in response to youdin's comments
fixed some typos, corrected kappa^2 dependence on dust
tiny correction in entropy def
head to toe edits

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  s = \ln{\left[K^\frac{1}{\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas perfectly-coupled to dust:  
the standard adiabatic energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation. 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
standard in many hydrodynamics codes.

@@

As noted in \S\ref{loc_iso_eos}, dust settling causes the effective
temperature to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, having a non-trivial vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this linear problem in a following 
study.

@@

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extrema in the quantity
\begin{align}
 \eta = \frac{\kappa^2}{2\Omega\Sigma}\mathcal{S}^{2/\gamma},  
\end{align} 
where $\mathcal{S} = \Pi/\Sigma^\gamma$. Here, $\Sigma$ is the total
surface density and $\Pi$ is the vertically integrated pressure.

From the dusty/adiabatic gas equivalence, we deduce the corresponding
condition for RWI in a (locally isothermal or polytroipc) dusty gas is
also given by extrema in $\eta$ with $\mathcal{S} = \exp{s}$ (and
$\tepsilon\to \Sigma_\mathrm{d}/\Sigma$). Since
the dusty entropy $s$ reflects the dust distribution, we expect that
RWI may also be triggered by extrema in  the dust-to-gas ratio.

@@

@@

Fig. \ref{compare_eigenvals_ts0d02} shows the unstable modes found for
this setup. We find the usual VSI modes, here with $\omega> 3\hgas\OmK,\, s<
0.2\hgas\OmK$. However, we also find modes with $\omega<s$. One of these,
marked by the blue asterisk, has 
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We visualize this mode in
Fig. \ref{result2d_loren}. We find this mode 
persists when $q=0$, so it is not the VSI. 
Using Eq. \ref{streaming_dispersion} with the above parameters,  
we also estimate a much smaller growth 
rate for the streaming instability of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}. However, the mode disappears with
$\tstop=0$, so it is associated with finite dust-gas drag.

@@

This preliminary calculation only serves to highlight the possibility
of new axisymmetric dust-drag instabilities in stratified disks. One 
issue is that the base state for the above calculation is not a true
equilibrium (although \citeauthor {loren15} find the dust settling
time is longer than the instability growth time). We will study
this instability in more detail with a proper basic
state in a follow up work.

\section{Summary}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas, and the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the dusty flow. This is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite dust-gas friction, $\tstop\neq 0$, the dust content of a
fluid parcel of the dusty mixture no longer conserved. A fluid parcel 
can exchange dust particles with neighboring parcels owing 
to dust-gas relative drift. This is analogous to heat exchange between
a  pure gas parcel and its surroundings.

We explicitly show that for a locally isothermal, dusty gas that the
evolutionary equation of  $\rhod/\rhog$ may be replaced by an 
effective energy equation. We find the 
natural definition of the entropy of the dusty gas is 
\begin{align*}
  s  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
In our framework, finite dust-gas friction appears as an energy
source term.  This analogy with standard adiabatic
hydrodynamics allow us to obtain dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

%
In this first study, we apply this equivalence principle to study the
axisymmetric stability of isothermal, protoplanetary disks
perfectly coupled to dust ($\tstop=0$). This is so that we may 
compute exact, steady equilibria for analyses. Furthermore, 
in this limit the dusty problem is \emph{identical} to adiabatic, pure
gas dynamics.

We obtain the Solberg-Hoiland criteria for axisymmetric stability for
strictly isothermal, perfectly-coupled dusty gas.  
We show that the vertical shear associated with thin dust
 layers \emph{cannot} lead to axisymmetric
  instabilities, however large it is.  
This is consistent with the fact that only 
\emph{non-axisymmetric} Kelvin-Helmholtz instabilities have been 
observed in numerical simulations in the perfectly-coupled limit 
\citep{chiang08,barranco09,lee10}. We show that 
sharp radial edges in the dust-to-gas ratio cannot persist in
protoplanetary disks, as these imply a sharp entropy gradient and may
thus destablize the disk.

%does not preclude non-axisymmetric instabilities. but if there is a
%global radial temp gradient, then vertical shear arising from
%it will always be comparable to dust-settling. but there is vsi with
%global rad temp grad

For locally isothermal disks, $c^2_s=c_s^2(r)$, we generalize the 
vertical shear instability (VSI) to dusty disks, which is still 
driven by the radial temperature gradient. We find that 
dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}.

%These results are readily
%understood by comparing the vertical shear rate with the buoyancy of
%an isothermal, dusty gas which we have correctly identified.

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. In particular, we emphasize the 
physical origin of any instability driven by strong dust-gas
drag is a phase lag between the pressure and total density evolution
of the dusty gas mixture. This leads to growing modes because 
positive work is done during an oscillation cycle, provided by
dust-gas drag, which is equivalent to a `heat' 
source. This interpretation is entirely analogous to pulsational
instabilities in stars.

>>>>

\subsection{Generalization to locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas perfectly-coupled to dust:  
the standard adiabatic energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
featured in many hydrodynamics codes.

@@

As noted in \S\ref{loc_iso_eos}, dust settling causes the effective
temperature to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, a non-trivial effective vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this linear problem in a following 
study.

@@

\subsubsection{Rossby wave instability}
The Rossby wave instability (RWI) is a non-axisymmetric, 2D shear
instability that operates in thin disks when it has radial structure
\citep{lovelace99,li00}. These studies consider adiabatic pure gas and 
show instability is possible if there is an extrema in the generalized
potential vorticity 
\begin{align}
 \mathcal{V} = \frac{\kappa^2}{2\Omega\Sigma}\mathcal{S}^{2/\gamma},  
\end{align} 
where $\mathcal{S} = \Pi/\Sigma^\gamma$. Here, $\Sigma$ is the total
surface density and $\Pi$ is the vertically integrated pressure.

From the dusty/adiabatic gas equivalence, we deduce the corresponding
condition for RWI in a (locally isothermal or polytroipc) dusty gas is
also given by extrema in $\mathcal{V}$ with $\mathcal{S} = \exp{\left(\Gamma
S_\mathrm{eff}\right)}$ and $\tepsilon = \Sigma_\mathrm{d}/\Sigma$. Since
the effective entropy reflects the dust distribution, we expect that
RWI may also be triggered by extrema in the dust-to-gas ratio, e.g. 
narrow dust rings. This would lead to direction formation of dusty
vortices.

%{\bf so may it's not vortex formation then dust-trapping. could just
%  form dusty vortices in one go}

@@

%{\bf vsi enabled by drag? slipping between dust and gas reduces
 % effective buoyancy}

@@

Fig. \ref{compare_eigenvals_ts0d02} shows the unstable modes found for
this setup. We find the usual VSI modes, here with $\omega> 3\hgas\OmK,\, s<
0.2\hgas\OmK$. However, we also find modes with $\omega<s$. One of these,
marked by the blue asterisk, has 
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We visualize this mode in
Fig. \ref{result2d_loren}. We find this mode 
persists when $q=0$, so it is not the VSI. 
Using Eq. \ref{streaming_dispersion} with the above parameters,  
we also estimate a much smaller growth 
rate for the streaming instability of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}. However, the mode disappears with $\tstop=0$, so it \emph{is}
associated with finite dust-gas drag.

@@

One possible interpretation of this mode is `drag-driven VSI', to be 
compared with `cooling-driven VSI' studied by \cite{lin15}. In the
latter case, buoyancy forces are eliminated by fast
cooling, allowing vertical shear to destabilize the disk. Now, in an
isothermal dusty disk if $\tstop=0$ then mixture is 
effectively adiabatic: dust-induced buoyancy forces are
strongest. However, having $\tstop\neq0$ allow dust particles to slip
past the gas, reducing buoyancy. That is, dust-gas drag
provies an effective cooling.

We stress that this example only serves to highlight
the possibility of new axisymmetric dust-drag instabilities in
stratified disks. A major caveat of the above result is that 
%the base state is not an exact equilibrium, because dust-settling would occur
%in practice. (
 the background dust settling timescale is comparable to the instability
growth timescale. It is essential to construct
exact equilibria for a meaningful stability analysis. We defer this
to future work.

%also unclear what are "proper" vertical bc for dusty disks

\section{Summary}\label{summary}
In this paper we describe the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas, and the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the dusty flow. This is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite dust-gas drag, $\tstop\neq 0$, the dust content of a 
parcel of the mixture no longer conserved. The parcel 
can exchange dust particles with neighboring parcels owing 
to dust-gas relative drift. This is analogous to heat exchange between
a pure gas parcel and its surroundings.

We explicitly show that for a locally isothermal, dusty gas that the 
evolutionary equation of  $\rhod/\rhog$ may be replaced by an 
effective energy equation. This leads to a 
natural definition of the effective entropy of isothermal dusty gas as  
\begin{align*}
  \seff  = \ln{\left(\frac{c_s^2\rhog}{\rhog+\rhod}\right)}.  
\end{align*}
%This allows us to define the buoyancy frequency of an isothermal 
%dusty gas. 
Finite dust-gas friction then appears as an energy
source term.  This analogy with standard 
hydrodynamics with cooling/heating allow us to obtain dusty analogs of gaseous
instabilities, and provide thermodynamical interpretations of  
dust-drag instabilities.

We obtain the Solberg-Hoiland criteria for axisymmetric stability for
strictly isothermal, perfectly-coupled dusty gas.  
Applying this to typical protoplanetary disks, we find that 
the vertical shear associated with dust 
 layers \emph{cannot} lead to axisymmetric  
  instabilities, however thin the dust layer is.   
%This is consistent with the fact that only 
%\emph{non-axisymmetric} Kelvin-Helmholtz instabilities have been 
%observed in numerical simulations in the perfectly-coupled limit 
%\citep{chiang08,barranco09,lee10}. 
Instead, sharp radial  edges in the dust-to-gas ratio could destabilized the
disk, as these imply sharp gradients in the disk's effective entropy
profile. % {\bf presumably this limits how sharp dust rings can be}

We  apply our thermodynamic framework to interpret
the streaming instability \citep[SI, ][]{youdin05a,jacquet11} and to generalize the vertical shear
instability \citep[VSI, ][]{nelson13,lin15} to dusty disks. We
explicitly show that the SI occurs because the evolution of gas
pressure lags behind  dust density; a general property of \emph{all}
instabilities driven by strong dust-gas drag. 
It takes a finite time for the gas to respond to
the dust motion. A lag implies a time interval when
gas pressure is still increasing while dust is already being expelled. 
The dusty gas then does positive work, which goes into amplifying
oscillation amplitudes. This interpretation is analogous to stellar pulsational
instabilities \citep{cox67}.

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}.

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. Because isothermal dusty gas has an
effective entropy, we suggest that purely hydrodynamic processes, such
as the Rossby Wave Instability \citep{li00} or the `Zombie Vortex
Instability' \citep{marcus15} could have dusty counter-parts.   
Furthermore, hydrodynamic instabilities driven by thermal cooling, such as VSI in
stably-stratified disks \citep{lin15} or the `Convective
Overstability' \citep{klahr14, lyra14},  may also find dusty analogs 
because finite dust-gas drag is equivalent to a heat sink/source.

%In particular, we emphasize the 
%physical origin of any instability driven by strong dust-gas
%drag is a phase lag between the pressure and total density evolution
%of the dusty gas mixture. This leads to growing modes because 
%positive work is done during an oscillation cycle, provided by
%dust-gas drag, which is equivalent to a `heat' 
%source.
